# Basic R functions
In this short chapter you will find a summary of some of the basic and most useful functions and operators from the tidyverse package that you will be using in R studio. If you want to learn a bit more about a specific function from this chapter simply click on the function, and it will take you to a page with additional information.

::: callout-note
## Functions

-   `library()` - Loads packages into R.
-   [`filter()`](basic_functions_extended.qmd#filter)- Allows you to retain/filter rows from a data set depending on the condition you want.
-   [`select()`](basic_functions_extended.qmd#select) - Allows you to pick/keep specific columns from a data set.
-   [`count()`](basic_functions_extended.qmd#count) - Counts the number of occurrences of all distinct elements in a column.
-   [`mutate()`](basic_functions_extended.qmd#mutate) - Allows you to create and modify columns from a data set.
-   [`arrange()`](basic_functions_extended.qmd#arrange) - Allows for the ordering of the data by different variables.
-   [`group_by()`](basic_functions_extended.qmd#group_by) - Uses an existing table to form a grouped table by splitting the data sate based on one or more variables.
-   [`summarise()`](basic_functions_extended.qmd#summarise) - Allows you to calculate summary statistics and is usually used on grouped data produced by `group_by()`, as it creates one row for each combination of grouping variables.
-   [`tibble()`](basic_functions_extended.qmd#tibble) - Allows you to create new data frames.
-   [`drop_na()`](basic_functions_extended.qmd#drop_na) - Allows you to to drop all the rows containing missing values.
:::

------------------------------------------------------------------------

## Joining functions

::: callout-note
## Functions

-   [`left_join()`](joining_functions_extended.qmd)- Keeps all the values from the left table and matching rows from the right table.
-   [`right_join()`](joining_functions_extended.qmd)- Keeps all the values from the right table and matching rows from the left table.
-   [`inner_join()`](joining_functions_extended.qmd)- Keeps all values that have matching keys in both tables.
-   [`full_join()`](joining_functions_extended.qmd)- Keeps all values from both tables.
:::

------------------------------------------------------------------------

## Pivot functions

::: callout-note
## Functions

-   [`pivot_longer()`](pivot_functions_extended.qmd)- Allows you to lengthen your data set from a wide format to a long format by reducing the number of columns and increasing the number of rows.
-   [`pivot_wider()`](pivot_functions_extended.qmd)- Allows you to widen your data set from a long format to a wide format by reducing the number of rows and increasing the number of columns.
:::

------------------------------------------------------------------------

## Useful operators

`|>` or `%>%`

The pipe operator lets you pass the result of one function directly into the next, reading left to right rather than from the inside out. This makes longer sequences of steps much easier to follow.

Without a pipe, applying several steps in sequence produces deeply nested code that is hard to read:

```{r}
#| include: false
library(tidyverse)
library(here)
rodents <- read_tsv(here("data", "rodents_full.txt"))
```

```{r}
#| eval: false
select(filter(rodents, species_id == "NL"), species_id, hindfoot_length, sex)
```

With the pipe, the same steps read as a plain sentence — "take rodents, then filter, then select":

```{r}
#| eval: false
rodents |>
  filter(species_id == "NL") |>
  select(species_id, hindfoot_length, sex)
```

::: callout-tip
## Keyboard shortcut

Type the pipe with **Ctrl + Shift + M** (Windows) or **Cmd + Shift + M** (macOS).
:::

::: callout-note
## `|>` vs `%>%` — what is the difference?

You will see both versions used online and in textbooks. `|>` is the **native pipe**, built into R since version 4.1 (2021) and requires no packages. `%>%` is the **magrittr pipe**, which comes from the `magrittr` package and is loaded automatically with `tidyverse`. They behave identically for the vast majority of everyday use, so do not worry about the difference for now. This book uses `|>`.
:::

::: callout-warning
## Common pipe errors

-   **`could not find function "%>%"`** — you are using `%>%` without loading `tidyverse` (or `magrittr`). Either add `library(tidyverse)` at the top of your script, or switch to `|>`.
-   **`unexpected symbol` or `object not found` after a pipe** — a line break in the wrong place. Make sure each pipe `|>` is at the *end* of a line, not the start of the next one.
:::

`%in%`

::: callout
-   This is a logical operator that allows you to check if an element from one vector is present in another vector.
:::

`!`

::: callout
-   This is a negation operator, which allows you to tell a function to do the opposite.
:::

This is an example of how you can oppose the filter() in order to filter **out** variables:

```{r}
#| eval: false
rodents |> 
  filter(!species_id=="NL")
```

`+`, `-`, `log()`, `*`, `/`

::: callout
-   These are all mathematical operators.
:::

------------------------------------------------------------------------

## Data overview functions

::: callout-note
## Functions

-   `class()` - It provides just the class of the specified object.
-   `str()` - It shows you the type of content in each column.
-   `glimpse()` - While still detailed, it is easier to glance through than str().
:::

------------------------------------------------------------------------

## Graph functions

`ggplot()` is the function that will allow you to plot your datasets.

Now, in order to get a plot you should think about three main components:

::: {style="text-align: center;"}
**Data + Aesthetics + Geometry**</span>
:::

::: callout
-   **Data** is a data frame you want to use (you can write it inside the ggplot() function of you can directly pipe ggplot() from your dataset.
-   **Aesthetics** is used to indicate x and y variables and it can also be used to control the color of the plot.
    -   `aes()`
-   **Geometry** refers to the type of plot you want to make:
    -   `geom_boxplot()`
    -   `geom_point()`
    -   `geom_line()`
    -   `geom_col()`
    -   `geom_histogram()`
    -   `geom_jitter()`
    -   `geom_bar()`
:::

Here you can see two examples of how to do a plot (one with pipe operator and one without):

```{r}
#| include: false  
library(tidyverse)
library(here) 
rodents <- read_tsv (here("data","rodents_full.txt"))
```

```{r}
#| warning: false
ggplot(rodents) + aes(x= species_id, y=hindfoot_length) + geom_boxplot()

rodents |> 
  ggplot() + aes(x= species_id, y=hindfoot_length) + geom_boxplot()
```

### Further Graph functions

::: callout-note
## Functions

-   `facet_wrap()` - Allows for the visualization of individual categories in their own graph.
-   `labs()` - Change axis labels and legend titles.
-   `scale_x_continuous() / scale_y_continuous()` - default scales for continuous x and y aesthetics, which can be used to transform these axes.
-   `theme()` - Modify a single plot's theme.
-   `theme_update()` - Modify the active theme, to affect all subsequent plots.
-   `theme_minimal()` - Applies a simplified theme of the plot, with no background shading.
:::

---

::: callout-note
## Chapter recap

**Functions covered:** `library()`, `filter()`, `select()`, `count()`, `mutate()`, `arrange()`, `group_by()`, `summarise()`, `drop_na()`, `tibble()`, `left_join()`, `right_join()`, `inner_join()`, `full_join()`, `pivot_longer()`, `pivot_wider()`, `ggplot()`, `aes()`, `geom_boxplot()`, `geom_point()` and others, `facet_wrap()`, `labs()`, `theme_minimal()`, `glimpse()`, `|>`, `%in%`, `!`

**Key idea:** The pipe `|>` is not just a shortcut — it changes the way you think about code. Instead of nesting functions inside each other, you write a sequence of steps in the order they happen, which makes the code much easier to read and debug.
:::
