# A bit more about basic functions

## Filter

The `filter()` function allows you to pick rows in a data set. To use this function you simply need to specify the requirement/condition by which you want to filter your data.

::: callout-tip
## You do this by writing:

filter(data set name, column name == condition you want to filter)
:::

In this example you can see both ways to use the `filter()` function in a data set:

```{r}
#| include: false  
library(tidyverse)
library(here)

rodents <- read_tsv(here("data", "rodents_full.txt"))
```

```{r}
#| eval: false
# Filter rodents data set for variables that are of the NL species.
filter(rodents, species_id=="NL")

rodents |> 
  filter(species_id=="NL")
```

### Useful filter functions and parameters

> -   `is.na()` - Allows you to check for missing values.
> -   `!is.na()` - Keeps only rows where a value is *not* missing (the `!` negates `is.na()`).
> -   `%in%` - Checks whether a value belongs to a list of options.
> -   `c()` - Combines multiple values into a list, used together with `%in%`.
> -   `na.rm = TRUE` - This parameter instructs functions to ignore missing values.

**Filtering for missing values**

`is.na()` returns `TRUE` for any cell that is empty (`NA`). Use it inside `filter()` to keep only rows where a column has a missing value, or use `!is.na()` to drop them:

```{r}
#| eval: false
# Keep only rows where hindfoot_length is missing
rodents |>
  filter(is.na(hindfoot_length))

# Keep only rows where hindfoot_length is NOT missing
rodents |>
  filter(!is.na(hindfoot_length))
```

**Filtering for multiple values with `%in%`**

Rather than chaining several `==` conditions with `|` (or), use `%in%` to check against a list of values in one step:

```{r}
#| eval: false
# Keep rows where species_id is one of three species
rodents |>
  filter(species_id %in% c("NL", "DM", "PF"))

# The equivalent using | is harder to read and easier to get wrong:
rodents |>
  filter(species_id == "NL" | species_id == "DM" | species_id == "PF")
```

You can also use `!` with `%in%` to *exclude* a set of values:

```{r}
#| eval: false
# Keep all rows except those three species
rodents |>
  filter(!species_id %in% c("NL", "DM", "PF"))
```

------------------------------------------------------------------------

## Select

The `select()` function allows you to pick/keep specific columns from a data set.

::: callout-tip
## You do this by writing:

select(data set name, column name/s)
:::

In this example you can see both ways to use the `select()` function in a data set:

```{r}
#| eval: false
# Select the species_id column from the rodents data set.
select(rodents, species_id)

rodents |> 
  select(species_id)
```

### Useful select functions

> -   `c()` - Allows you to combine different selections.
> -   `starts_with()` - Allows you to select variables that start with a specific prefix.
> -   `ends_with()` - Allows you to select variables that end with a specific suffix
> -   `contains()` - Allows you to select variables that contain a specific string.

------------------------------------------------------------------------

## Count

The `count()` function counts the number of occurrences of all distinct elements in a column.

::: callout-tip
## You do this by writing:

count(data set name, column name)
:::

In this example you can see both ways to use the `count()` function in a data set:

```{r}
#| eval: false
# Select the species_id column from the rodents data set.
count(rodents, species_id)

rodents |> 
  count(species_id)
```

### Useful count parameters and functions

> -   `sort = TRUE/FALSE` - This parameters dictates the order in which the variables will be arranged, TRUE will order the counts in a descending order, wile FALSE will order the counts in an ascending order.

------------------------------------------------------------------------

## Mutate

The `mutate()` function allows you to create and modify columns from a data set. - It can create new columns as functions of existing variables.

-   It can modify existing columns.
-   It can delete columns by setting them to NULL.

::: callout-tip
## You do this by writing:

mutate(data set name, new/existing column name = function/expression)
:::

### Useful mutate functions

> -   `+`, `-`, `log()`, `*`, `/` - Mathematical operators
> -   `ifelse()` - This function will return one value if a condition is TRUE and another if FALSE.
> -   `case_when()` - This function allows you to evaluate multiple conditions and assign them individual values.

In this examples you can see different ways you can use the `mutate()` function in a data set:

```{r}
#| eval: false
# Use mutate to create a new column that will show a 1 for females and a 2 for males
rodents |> 
  mutate(new_column = case_when(sex == "F" ~ 1, sex == "M" ~ 2))

# Use mutate to create a new weight_kg column that converts the values on the weight column (in grams) into kilograms by dividing them
rodents |> 
  mutate(wight_kg=weight/1000)

# Use mutate to set the delete the hindfoot_length column
rodents |> 
  mutate(hindfoot_length = NULL)

```

------------------------------------------------------------------------

## Arrange

The `arrange()` function allows for the ordering of the data by different variables. - If you input multiple columns they will be sorted hierarchically.

::: callout-tip
## You do this by writing:

arrange(data set name, column name/s)
:::

In this example you can see both ways to use the `arrange()` function in a data set:

```{r}
#| eval: false
#Arrange the rows so that they sorted by species_id (ascending) and then by sex (ascending) 
arrange(rodents, species_id, sex)

rodents |> 
  arrange(species_id, sex)
```

### Useful arrange functions

> `desc()` - Allows you to sort in descending order as the default order of the `arrange()` funcrion is ascending.

```{r}
#| eval: false
rodents |> 
  arrange(desc(year))
```

------------------------------------------------------------------------

## Group_by

The `group_by()` function uses an existing table to form a grouped table by splitting the data sate based on one or more variables.

::: callout-tip
## You do this by writing:

group_by(data set name, column name/s)
:::

In this example you can see both ways to use the `group_by()` function in a data set:

```{r}
#| eval: false
# Group the data set by species_id
group_by(rodents, species_id)

rodents |> 
  group_by(species_id)
```

------------------------------------------------------------------------

## Summarise

The `summarise()` function allows you to calculate summary statistics and is usually used on grouped data produced by `group_by()`, as it creates one row for each combination of grouping variables.

::: callout-tip
## You do this by writing:

summarise(data set name, new column name = function(column name))
:::

### Useful summary functions and parameters

-   `mean()`
-   `median()`
-   `sd()`
-   `min()`
-   `max()`
-   `n()`
-   `n_distinct()`
-   `na.rm = TRUE` - This parameter instructs functions to ignore missing values.

In this example you can see both ways to use the `summarise()` function in a data set:

```{r}
#| eval: false
# Take the mean hindfoot length
summarise(rodents, mean_hindfoot_length = mean(hindfoot_length, na.rm = TRUE))

rodents |> 
  summarise(mean_hindfoot_length = mean(hindfoot_length, na.rm = TRUE))
```

------------------------------------------------------------------------

## Tibble

The `tibble()` function allows you to create new data frames.

::: callout-tip
## You do this by writing:

tibble(data set name, column name 1 = variables, comuln name 2 = variables, ...)
:::

### Useful tibble functions

> -   `c()` - Allows you to combine different selections.

```{r}
tibble(
  Teams = c("Blue", "Red", "Purple"),
  Score = c(100, 85, 99))
```

------------------------------------------------------------------------

## Drop_na

The `drop_na()` function allows you to to drop all the rows containing missing values.

::: callout-tip
## You do this by writing:

drop_na(data set name, column name)
:::

In this example you can see both ways to use the `drop_na()` function in a data set:

```{r}
#| eval: false
# Drop all the variables that have missing values in the hindfoot length column
drop_na(rodents, hindfoot_length)

rodents |>
  drop_na(hindfoot_length)
```

---

::: callout-note
## Chapter recap

**Functions covered:** `filter()`, `select()`, `count()`, `mutate()`, `arrange()`, `group_by()`, `summarise()`, `drop_na()`, `tibble()`

**Key parameters and helpers:** `is.na()`, `!is.na()`, `%in%`, `c()`, `desc()`, `sort = TRUE`, `na.rm = TRUE`, `ifelse()`, `case_when()`, `starts_with()`, `ends_with()`, `contains()`

**Key idea:** Almost every tidyverse function follows the same pattern — first argument is the data, rest describes what to do. That consistency is why they all slot together so naturally with the pipe.
:::
