# Graph Functions

`ggplot2` builds plots in layers. You start with the data, then add aesthetics (which columns map to which visual properties), then a geometry (the type of plot), and optionally labels, scales, and a theme. Each layer is added with `+`.

```{r}
#| include: false
library(tidyverse)
library(here)
rodents <- read_tsv(here("data", "rodents_full.txt"))
```

## Building a plot layer by layer

It helps to see a plot grow one layer at a time.

**Step 1 — data only:** `ggplot()` sets up the coordinate system but draws nothing yet.

```{r}
ggplot(data = rodents)
```

**Step 2 — add aesthetics:** `aes()` tells ggplot which columns to use for x and y. Still nothing drawn.

```{r}
ggplot(data = rodents, aes(x = species_id, y = hindfoot_length))
```

**Step 3 — add a geometry:** `geom_boxplot()` draws the plot.

```{r}
#| warning: false
ggplot(data = rodents, aes(x = species_id, y = hindfoot_length)) +
  geom_boxplot()
```

**Step 4 — add labels:** `labs()` replaces the default column-name labels.

```{r}
#| warning: false
ggplot(data = rodents, aes(x = species_id, y = hindfoot_length)) +
  geom_boxplot() +
  labs(x = "Species", y = "Hindfoot length (mm)", title = "Hindfoot length by species")
```

**Step 5 — apply a theme:** `theme_minimal()` removes the grey background for a cleaner look.

```{r}
#| warning: false
ggplot(data = rodents, aes(x = species_id, y = hindfoot_length)) +
  geom_boxplot() +
  labs(x = "Species", y = "Hindfoot length (mm)", title = "Hindfoot length by species") +
  theme_minimal()
```

## Colour

Map a column to `colour` (for lines/points) or `fill` (for bars/boxes) inside `aes()` to colour by group:

```{r}
#| warning: false
ggplot(data = rodents, aes(x = species_id, y = hindfoot_length, fill = sex)) +
  geom_boxplot() +
  labs(x = "Species", y = "Hindfoot length (mm)", fill = "Sex") +
  theme_minimal()
```

## Faceting

Faceting splits one plot into a grid of panels, one per group. It is often clearer than colouring by group when there are many categories. Use `facet_wrap()` with the column you want to split on, preceded by `~`:

```{r}
#| warning: false
ggplot(data = rodents, aes(x = hindfoot_length)) +
  geom_histogram(bins = 20) +
  facet_wrap(~ sex) +
  labs(x = "Hindfoot length (mm)", y = "Count") +
  theme_minimal()
```

::: callout-tip
## When to use faceting vs colour

Use **colour** when you have two or three groups and want to compare them directly on the same axes.

Use **`facet_wrap()`** when you have more groups, or when the groups overlap so much that colours become hard to read.
:::

## Common geometries

| Function | Plot type |
|---|---|
| `geom_point()` | Scatter plot |
| `geom_line()` | Line plot |
| `geom_boxplot()` | Box plot |
| `geom_histogram()` | Histogram |
| `geom_bar()` | Bar chart (counts rows) |
| `geom_col()` | Bar chart (uses a value column for height) |
| `geom_jitter()` | Scatter with random horizontal spread (useful for discrete x) |

---

::: callout-note
## Chapter recap

**Functions covered:** `ggplot()`, `aes()`, `geom_point()`, `geom_line()`, `geom_boxplot()`, `geom_histogram()`, `geom_bar()`, `geom_col()`, `geom_jitter()`, `labs()`, `facet_wrap()`, `theme_minimal()`

**Key parameters:** `colour`, `fill` (inside `aes()` to colour by a variable); `bins` (for `geom_histogram()`); `~ variable` (for `facet_wrap()`)

**Key idea:** Every ggplot is three things — data, what to map where (`aes()`), and what shape to draw (`geom_*()`). Everything else — labels, colours, themes, facets — is optional polish you add one `+` at a time.
:::
