# Joining functions

The joining functions allow you to merge separate data sets based on a shared **key** column — a column that appears in both tables and contains matching values, such as a species ID or a sample name.

## The four join types

| Function | Rows kept | Use when… |
|---|---|---|
| `left_join()` | All rows from the **left** table; matching rows from the right | You want to keep all your primary data and add extra columns where available |
| `right_join()` | All rows from the **right** table; matching rows from the left | Same idea, but your primary data is on the right |
| `inner_join()` | Only rows with a matching key **in both** tables | You only want rows that exist in both data sets |
| `full_join()` | All rows from **both** tables | You want to keep everything and fill gaps with `NA` |

Rows with no match in the other table get `NA` in the columns that came from that table.

## Seeing joins in action

The examples below use two small built-in dplyr data sets. Here is what they contain:

```{r}
#| echo: false
library(tidyverse)
```

```{r}
band_members      # name + band
band_instruments  # name + plays
```

`Mick` is in `band_members` but not in `band_instruments`. `Keith` is in `band_instruments` but not in `band_members`. The key column is `name`.

```{r}
# inner_join: only Mick and John appear in both tables — only they are kept
inner_join(band_members, band_instruments, by = "name")

# left_join: all of band_members is kept; Keith (right only) is dropped
left_join(band_members, band_instruments, by = "name")

# right_join: all of band_instruments is kept; Mick (left only) is dropped
right_join(band_members, band_instruments, by = "name")

# full_join: everyone is kept; gaps filled with NA
full_join(band_members, band_instruments, by = "name")
```

## Joining on columns with different names

If the key column has a different name in each table, use `join_by()` to specify which columns match:

```{r}
#| eval: false
# key is called "name" on the left and "artist" on the right
left_join(band_members, band_instruments2, by = join_by(name == artist))
```

::: callout-tip
## You do this by writing:

joining_function(left_table, right_table, by = "shared_column_name")

\# or, if the key columns have different names:

joining_function(left_table, right_table, by = join_by(left_col == right_col))
:::
