# Even more about pivot functions

The pivot functions allow you to reshape your data sets, which is useful when preparing your data for analysis or visualization.

-   `pivot_longer()` - Converts **wide** data to **long** format: fewer columns, more rows.
-   `pivot_wider()` - Converts **long** data to **wide** format: more columns, fewer rows.

## Wide vs long format

It helps to see what these words mean before looking at code. Imagine a small table recording hindfoot length and weight for two animals:

**Wide format** (one row per animal, measurements spread across columns):

| species_id | hindfoot_length | weight |
|------------|-----------------|--------|
| NL         | 32              | 165    |
| DM         | 36              | 43     |

**Long format** (one row per measurement, a column for the measurement name and a column for the value):

| species_id | type_measurement | measurements |
|------------|------------------|--------------|
| NL         | hindfoot_length  | 32           |
| NL         | weight           | 165          |
| DM         | hindfoot_length  | 36           |
| DM         | weight           | 43           |

::: callout-note
## When to use each

-   Use **`pivot_longer()`** (wide → long) when you want to plot or summarise multiple measurements together — for example, drawing a boxplot of several numeric columns at once with `ggplot2`.
-   Use **`pivot_wider()`** (long → wide) when you need one column per group for comparison — for example, to create a summary table where each species gets its own column.
:::

## `pivot_longer()`

::: callout-tip
## You do this by writing:

pivot_longer(data, cols = first_col:last_col, names_to = "new_name_column", values_to = "new_value_column")
:::

-   `cols` - The columns you want to pivot (use an unquoted range like `hindfoot_length:weight`).
-   `names_to` - The name of the new column that will hold the old column names.
-   `values_to` - The name of the new column that will hold the values.

```{r}
#| eval: false
rodents |>
  pivot_longer(cols = hindfoot_length:weight, names_to = "type_measurement", values_to = "measurements")
```

## `pivot_wider()`

::: callout-tip
## You do this by writing:

pivot_wider(data, names_from = column_with_names, values_from = column_with_values)
:::

-   `names_from` - The column whose values will become the new column names.
-   `values_from` - The column whose values will fill those new columns.

```{r}
#| eval: false
# Example: count captures per species per year, then spread years into columns
rodents |>
  count(species_id, year) |>
  pivot_wider(names_from = year, values_from = n)
```

---

::: callout-note
## Chapter recap

**Functions covered:** `pivot_longer()`, `pivot_wider()`

**Key parameters:** `cols`, `names_to`, `values_to` (for `pivot_longer`); `names_from`, `values_from` (for `pivot_wider`)

**Key idea:** ggplot and most tidyverse analysis functions want data in long format — one row per observation, one column per variable. If your data is spread across many columns, `pivot_longer()` is almost always the first step.
:::
